---
title: "Auditory Oddball – Data Quality Report"
author: "Data Quality"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# REQUIRED PACKAGES
pkgs <- c("tidyverse", "data.table")

# Install packages if not installed
installed_packages = pkgs %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(pkgs[!installed_packages])
}

# Load packages
library(tidyverse)
library(data.table)

# Display Session Info
sessionInfo()
```

## Introduction
This report documents the data quality steps for the Auditory Oddball task. The pipeline includes:
1. PD NA check & summary
2. Exclusion of trials with > 50% PD NAs
3. Gaze offset check & exclusion (AOI threshold = 200 px from center)

```{r load data}
# Paths – adjust to your project
home_path <- "S:/KJP_Studien"
# home_path <- "//192.168.88.212/daten/KJP_Studien" # Alternative path
data_path <- "/LOCUS_MENTAL/6_Versuchsdaten/auditory_oddball/"

# Load data
df_et <- readRDS(paste0(home_path, data_path, "eyetracking_ao.rds"))
```

### Step 1: PD NA Check & Summary
We first calculate the proportion of missing Pupil Dilation (PD) data per trial and per participant.

```{r pressure, echo=FALSE}
# Per-trial NA summary
trial_na_summary <- df_et %>%
  group_by(id, trial_number) %>%
  summarise(
    total_samples = n(),
    na_count      = sum(is.na(pd)),
    na_prop       = mean(is.na(pd)),
    .groups = "drop"
  )

# Overall summary
summary(trial_na_summary$na_prop)

# Per-participant NA summary
participant_na_summary <- df_et %>%
  group_by(id) %>%
  summarise(
    total_samples = n(),
    na_count      = sum(is.na(pd)),
    na_prop       = mean(is.na(pd)),
    .groups = "drop"
  ) %>%
  arrange(desc(na_prop))

# View top participants with missing data
head(participant_na_summary, 10)
```

### Step 2: Trial Exclusion (High NA)
Trials with more than 50% missing pupil dilation data are excluded from further analysis.
```{r pd-exclusion}
na_threshold <- 0.50

high_na_trials <- trial_na_summary %>%
  filter(na_prop > na_threshold)

# Exclude flagged trials from the eye-tracking data
df <- df_et %>%
  anti_join(high_na_trials, by = c("id", "trial_number"))

# Exclusion summary statistics
total_trials     <- nrow(trial_na_summary)
excluded_trials  <- nrow(high_na_trials)
remaining_trials <- total_trials - excluded_trials

# Results summary
exclusion_stats <- data.frame(
  Metric = c("Total trials", "Excluded (NA > 50%)", "Remaining"),
  Count = c(total_trials, excluded_trials, remaining_trials),
  Percentage = c(100, (excluded_trials/total_trials)*100, (remaining_trials/total_trials)*100)
)

knitr::kable(exclusion_stats, digits = 2, caption = "Summary of NA filtering")

```

```{r na-comparison}
# Mean NA proportion in excluded vs. included trials
included_na_summary <- trial_na_summary %>%
  anti_join(high_na_trials, by = c("id", "trial_number"))

cat("Mean NA prop in excluded trials:", mean(high_na_trials$na_prop), "\n")
cat("Mean NA prop in included trials:", mean(included_na_summary$na_prop), "\n")
```

### Step 3: Gaze Offset Check
We classify samples based on their distance from the screen center (AOI = ±200 px).
#### 3a. Gaze Pre-processing
We average the left and right eye coordinates to get a single gaze point.
```{r gaze processing}
# Check for required columns
gaze_cols <- c("left_gaze_x", "right_gaze_x", "left_gaze_y", "right_gaze_y")
missing_gaze_cols <- setdiff(gaze_cols, names(df))

if (length(missing_gaze_cols) > 0) {
  stop("Missing gaze columns: ", paste(missing_gaze_cols, collapse = ", "))
}

# Calculate averages
df <- df %>%
  mutate(
    gaze_x_px = (left_gaze_x + right_gaze_x) / 2,
    gaze_y_px = (left_gaze_y + right_gaze_y) / 2
  )

# Summaries
summary(df %>% select(gaze_x_px, gaze_y_px))
```

#### 3b. AOI Classification
Samples are flagged if they fall outside the 200px boundary.

```{r aoi classification}
aoi_threshold <- 200

df <- df %>%
  mutate(
    aoi_label      = ifelse(
      abs(gaze_x_px) <= aoi_threshold & abs(gaze_y_px) <= aoi_threshold,
      "AOI", "Offset"
    ),
    gaze_distance  = sqrt(gaze_x_px^2 + gaze_y_px^2),
    gaze_deviated  = gaze_distance > aoi_threshold
  )

# Frequency table
table(df$aoi_label)

```

#### 3c. Gaze Exclusion Summary
```{r gaze-exclusion}
total_samples_before <- nrow(df)
df_aoi               <- df %>% filter(aoi_label == "AOI")
excluded_samples     <- total_samples_before - nrow(df_aoi)

cat(sprintf("Excluded Samples: %d (%.2f%%)\n", excluded_samples, (excluded_samples/total_samples_before)*100))
cat(sprintf("Remaining Samples: %d (%.2f%%)\n", nrow(df_aoi), (nrow(df_aoi)/total_samples_before)*100))

```

Final Data Summary

```{r final-summary}
# Per-trial offset summary for the final dataset
trial_offset_summary <- df %>%
  group_by(id, trial_number) %>%
  summarise(
    total_samples   = n(),
    offset_samples  = sum(aoi_label == "Offset", na.rm = TRUE),
    offset_prop     = mean(aoi_label == "Offset", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(offset_prop))

cat("Final Clean Dataset Statistics:\n")
cat("Rows: ", nrow(df_aoi), "\n")
cat("Participants: ", length(unique(df_aoi$id)), "\n")

```

