---
title: "Auditory Oddball Data Quality and Analysis"
author: "Iskra Todorova"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true               # shows table of contents (agenda)
      toc_depth: 3            # how many heading levels to include
      toc_float: true         # makes the TOC float on the side
      code_folding: hide      # allows collapsing code blocks (default: shown)
      number_sections: true   # optional, section numbers
      theme: flatly           # optional Bootstrap theme (e.g., cerulean, journal)
      highlight: tango        # syntax highlight style
---

```{r setup,echo = FALSE, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

pkgs <- c("tidyverse",
         "ggplot2", # creating graphs
         "dplyr",
         "patchwork",
         "knitr",
         "viridis",
         "DT",
         "kableExtra",
         "lme4",
         "lmerTest",
         "emmeans",
         "data.table",
         "hexbin"
          )

# check if required packages are installed
installed_packages = pkgs %in% rownames(installed.packages())

# install packages if not installed
if (any(installed_packages == FALSE)) {
  install.packages(pkgs[!installed_packages])
}

#load packages
lapply(pkgs, function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    message(paste("Package", pkg, "not found."))
  }
})
```

```{r load_data, include=FALSE, warning=FALSE}
# Define paths (adjust to your project)

#home_path <- "S:/KJP_Studien"
home_path <- "//192.168.88.212/daten/KJP_Studien"
data_path <- "/LOCUS_MENTAL/6_Versuchsdaten/auditory_oddball/"

# Load processed files
df_et <- readRDS(paste0(home_path, data_path, "eyetracking_ao.rds")) #uncleaned as created with 01_preprocessing
df_trial <- readRDS(paste0(home_path, data_path, "trialdata_ao.rds")) #uncleaned as created with 01_preprocessing

# #correct df_trial for cleaned data
# trial_meta <- unique(df_et[, .(id, trial_number)])
# df_trial<-merge(df_trial,trial_meta,by=c("id","trial_number"))
# ####--> now same data base of df_trial and  df_et

```

# Trial-Level Metrics
## Trial Duration, Stimulus Duration and ISI Duration

```{r trial_quality_plots_2, echo=TRUE, fig.width=12, fig.height=8}

df_trial <- df_trial[, trial_duration := ISI_end_time - stimulus_start_time]

# Expected durations
expected_trial_duration <- 2.2 # e.g., 1.8 to 2s ISI + ~150ms Stimulus
expected_stimulus_duration <- 0.15
expected_cartoon_duration <- 2

# Plot 1: Trial Duration
p1 <- ggplot(df_trial, aes(x = trial_duration)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~trial)+
  geom_vline(xintercept = expected_trial_duration, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = expected_trial_duration, y = 0, label = "Expected Max Duration",
           color = "red", angle = 0, vjust = -0.7, size = 3) +
  labs(title = "Distribution of Trial Duration", x = "Trial Duration (s)", y = "Count") +
  theme_minimal()

# Plot 2: Stimulus Duration
p2 <- ggplot(df_trial, aes(x = stimulus_duration)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~trial)+
  geom_vline(xintercept = expected_stimulus_duration, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = expected_stimulus_duration, y = 0, label = "Expected Duration",
           color = "red", angle = 0, vjust = -0.7, size = 3) +
  labs(title = "Distribution of Stimulus Duration", x = "Stimulus Duration (s)", y = "Count") +
  theme_minimal()

# Plot 3: ISI Duration
p3 <- ggplot(df_trial, aes(x = ISI_duration)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~trial)+
  geom_vline(xintercept = expected_cartoon_duration, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = expected_cartoon_duration, y = 0, label = "Expected Max Duration",
           color = "red",angle = 0, vjust = -0.7, size = 3) +
  labs(title = "Distribution of ISI Duration", x = "ISI Duration (s)", y = "Count") +
  theme_minimal()

# Print plots
p1 /p2 /p3 

```

## filter trials with excessive duration

```{r filter_trials}
# Filter trials longer than 2.5 seconds
long_trials <- df_trial %>%
  filter(trial_duration > 2.5)

df_t_filtered<- df_trial %>%
  anti_join(long_trials, by = c("id", "trial_number"))

# Step 3: Summary statistics
total_trials <- nrow(df_trial)
excluded_trials <- nrow(long_trials)
remaining_trials <- nrow(df_t_filtered)

percent_excluded <- (excluded_trials / total_trials) * 100
percent_remaining <- (remaining_trials / total_trials) * 100

cat(
  "Trial duration outlier filtering summary:\n",
  sprintf("Total trials: %d\n", total_trials),
  sprintf("Excluded trials: %d (%.2f%%)\n", excluded_trials, percent_excluded),
  sprintf("Remaining trials: %d (%.2f%%)\n", remaining_trials, percent_remaining),

  sprintf("Mean trial_duration of excluded trials: %.3f\n", mean(long_trials$trial_duration, na.rm = TRUE)),
  sprintf("Mean trial_duration of remaining trials: %.3f\n", mean(df_t_filtered$trial_duration, na.rm = TRUE))
  )

df_trial <- df_t_filtered
```
# Pupil Dilation (PD) Data 

## pupil data missing visualization

```{r pupil_quality_plot, echo=TRUE, fig.width=12, fig.height=6}
na_threshold <- 0.5 

if("pd" %in% names(df_et)) {
  # Plot pupil size over time with missing data indicators
  participant_data <- df_et %>% 
    mutate(sample_index = row_number())
  
  p1 <- ggplot(participant_data, aes(x = sample_index, y = pd)) +
    geom_line(color = "black", na.rm = TRUE, alpha = 0.7) +
    geom_point(data = subset(participant_data, is.na(pd)), 
               aes(x = sample_index, y = 0), 
               color = "red", shape = 4, size = 1) +
    labs(
      title = "Pupil Size Over Time (All Participants)",
      x = "Sample Index",
      y = "Pupil Diameter (mm)"
    ) +
    theme_minimal()
  
  # Distribution of missing data by participant
  missing_by_participant <- df_et%>%
    group_by(id) %>%
    summarise(
      missing_prop = mean(is.na(pd)),
      .groups = "drop"
    )
  
  p2 <- ggplot(missing_by_participant, aes(x = reorder(id, missing_prop), y = missing_prop)) +
    geom_col(fill = "steelblue") +
    labs(
      title = "Missing PD Data by Participant",
      x = "Participant ID",
      y = "Proportion Missing"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    geom_hline(yintercept = na_threshold, color = "red", linetype = "dashed")
  
  p1 / p2
}
```

## Missing PD data

Exploration of the missing pupil data across participants and trials. We examined the the proportion of missing PD data in the eye-tracking data set. We calculated the percentage of missing PD data for each trial and participant. We consider a 50 % and above missing data proportion in a trial as low quality data. Trials with missing data over 50% should be excluded.

```{r, echo=FALSE}

# Per-trial NA summary
trial_na_summary <- df_et %>%
  group_by(id, trial_number) %>%
  summarise(
    total_samples = n(),
    na_count      = sum(is.na(pd)),
    na_prop       = mean(is.na(pd)),
    .groups = "drop"
  )

table(trial_na_summary$na_prop)

missing_trials<-as.numeric(sum(trial_na_summary$na_prop))
total_trials<-as.numeric(nrow(trial_na_summary))

paste0('trials with missing data: ',missing_trials)
paste0('of total trials: ',total_trials)
paste0('trials with missing data percentage: ',round(missing_trials/total_trials,2))

# Per-participant NA summary
participant_na_summary <- df_et %>%
  group_by(id) %>%
  summarise(
    total_samples = n(),
    na_count      = sum(is.na(pd)),
    na_prop       = mean(is.na(pd)),
    .groups = "drop"
  ) %>%
  arrange(desc(na_prop))

cat("\nPer-participant PD missing data proportion:\n")
print(participant_na_summary)
sort(round(participant_na_summary$na_prop,2))

### MArch 2026:
#--> of n=91 datasets, n=84 have less than 10% missing data
```
## exclude trials with missing

```{r exclude NA trials, echo=FALSE}

na_threshold <- 0.50

high_na_trials <- trial_na_summary %>%
  filter(na_prop > na_threshold)

# Exclude flagged trials from the eye-tracking data
df <- df_et %>%
  anti_join(high_na_trials, by = c("id", "trial_number"))

# Exclusion summary
total_trials     <- nrow(trial_na_summary)
excluded_trials  <- nrow(high_na_trials)
remaining_trials <- total_trials - excluded_trials

cat(sprintf(
  "\nNA filtering summary:
  Total trials         : %d
  Excluded (NA > %.0f%%): %d  (%.2f%%)
  Remaining            : %d  (%.2f%%)\n",
  total_trials,
  na_threshold * 100,
  excluded_trials,  (excluded_trials  / total_trials) * 100,
  remaining_trials, (remaining_trials / total_trials) * 100
))

# Mean NA proportion in excluded vs. included trials
included_na_summary <- trial_na_summary %>%
  anti_join(high_na_trials, by = c("id", "trial_number"))

cat(sprintf(
  "  Mean NA prop in excluded trials : %.3f
  Mean NA prop in included trials  : %.3f\n\n",
  mean(high_na_trials$na_prop),
  mean(included_na_summary$na_prop)
))

```
# Gaze Position 

```{r gaze_preparation, echo=TRUE}
# Check if gaze columns exist
gaze_cols <- c("left_gaze_x", "right_gaze_x", "left_gaze_y", "right_gaze_y")
missing_gaze_cols <- setdiff(gaze_cols, names(df))

if(length(missing_gaze_cols) > 0) {
  cat("Warning: Missing gaze columns:", paste(missing_gaze_cols, collapse = ", "), "\n")
  cat("Available columns:", paste(names(df), collapse = ", "), "\n")
} else {
  # Calculate average gaze positions
  df$gaze_x_px <- (df$left_gaze_x + df$right_gaze_x) / 2
  df$gaze_y_px <- (df$left_gaze_y + df$right_gaze_y) / 2
  
  # Screen parameters
  monitor_width <- 2560
  monitor_height <- 1440
  half_width <- monitor_width / 2
  half_height <- monitor_height / 2
  
  cat("Gaze position summaries:\n")
  cat("X position: ")
  print(summary(df$gaze_x_px))
  cat("Y position: ")
  print(summary(df$gaze_y_px))
}
```

```{r gaze_plots, echo=TRUE, fig.width=12, fig.height=8}
# Function for plotting gaze with monitor boundaries
plot_gaze_with_boundaries <- function(data, half_width, half_height) {
  # Classify points as on/off monitor
  data$on_monitor <- abs(data$gaze_x_px) <= half_width & abs(data$gaze_y_px) <= half_height
  
  ggplot(data, aes(x = gaze_x_px, y = gaze_y_px)) +
    geom_point(aes(color = on_monitor), alpha = 0.3, size = 0.5) +
    scale_color_manual(values = c("FALSE" = "red", "TRUE" = "blue"),
                       labels = c("Off Monitor", "On Monitor")) +
    geom_rect(aes(xmin = -half_width, xmax = half_width,
                  ymin = -half_height, ymax = half_height),
              fill = NA, color = "green", linetype = "dashed", size = 1) +
    geom_vline(xintercept = 0, color = "black", linetype = "dotted") +
    geom_hline(yintercept = 0, color = "black", linetype = "dotted") +
    coord_fixed() +
    theme_minimal() +
    labs(
      title = "Gaze Points with Monitor Boundaries",
      x = "Gaze X (pixels)",
      y = "Gaze Y (pixels)",
      color = "Position"
    ) +
    theme(legend.position = "right")
}

if(length(missing_gaze_cols) == 0) {
  # Remove extreme outliers for better visualization
  df_plot <- df %>%
    filter(!is.na(gaze_x_px) & !is.na(gaze_y_px)) %>%
    filter(abs(gaze_x_px) < 3000 & abs(gaze_y_px) < 3000)  # Remove extreme outliers
  
  # Raw gaze distribution
  p1 <- ggplot(df_plot, aes(x = gaze_x_px, y = gaze_y_px)) +
    geom_hex(bins = 50) +
    scale_fill_viridis_c(option = "plasma", trans = "log10") +
    theme_minimal() +
    labs(
      title = "Gaze Position Distribution (Hex Plot)",
      x = "Gaze X (pixels)",
      y = "Gaze Y (pixels)",
      fill = "Log10 Count"
    ) +
    geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed")
  
  # Monitor boundaries plot
  p2 <- plot_gaze_with_boundaries(df_plot, half_width, half_height)
  
  p1 / p2
}

```

## remove deviating gaze data

outside an 200px AOI

```{r exclude_offmonitor_gaze, echo=TRUE}
cat("==========================================================\n")
cat("STEP 3 – Gaze Offset Check & Exclusion (AOI ±200 px)\n")
cat("==========================================================\n\n")

# --- 3a. Verify required gaze columns ----------------------------------------
gaze_cols         <- c("left_gaze_x", "right_gaze_x", "left_gaze_y", "right_gaze_y")
missing_gaze_cols <- setdiff(gaze_cols, names(df))

if (length(missing_gaze_cols) > 0) {
  stop("Missing gaze columns: ", paste(missing_gaze_cols, collapse = ", "))
}

# --- 3b. Average left/right gaze to get a single x/y per sample --------------
df <- df %>%
  mutate(
    gaze_x_px = (left_gaze_x + right_gaze_x) / 2,
    gaze_y_px = (left_gaze_y + right_gaze_y) / 2
  )

cat("Gaze position summaries (averaged eye):\n")
cat("  X position:\n"); print(summary(df$gaze_x_px))
cat("  Y position:\n"); print(summary(df$gaze_y_px))

# --- 3c. AOI classification --------------------------------------------------
# Screen centre is (0, 0) in gaze-coordinate space; AOI = ±200 px square
aoi_threshold <- 200

df <- df %>%
  mutate(
    aoi_label      = ifelse(
      abs(gaze_x_px) <= aoi_threshold & abs(gaze_y_px) <= aoi_threshold,
      "AOI", "Offset"
    ),
    gaze_distance  = sqrt(gaze_x_px^2 + gaze_y_px^2),   # Euclidean distance from centre
    gaze_deviated  = gaze_distance > aoi_threshold        # circular boundary flag
  )

aoi_counts <- table(df$aoi_label)
cat("\nGaze classification (box criterion ±200 px):\n")
print(aoi_counts)
cat("\nProportion (%):\n")
print(round(prop.table(aoi_counts) * 100, 2))

prop_deviated <- mean(df$gaze_deviated, na.rm = TRUE) * 100
cat(sprintf(
  "\nProportion of samples outside circular AOI boundary (r > %d px): %.2f%%\n",
  aoi_threshold, prop_deviated
))

# --- 3d. Keep only AOI samples -----------------------------------------------
total_samples_before <- nrow(df)
df_aoi               <- df %>% filter(aoi_label == "AOI")
excluded_samples     <- total_samples_before - nrow(df_aoi)

cat(sprintf(
  "\nGaze offset exclusion summary:
  Total samples         : %d
  Excluded (Offset)     : %d  (%.2f%%)
  Remaining (AOI only)  : %d  (%.2f%%)\n\n",
  total_samples_before,
  excluded_samples,      (excluded_samples      / total_samples_before) * 100,
  nrow(df_aoi),          (nrow(df_aoi)           / total_samples_before) * 100
))

# --- 3e. Per-trial offset summary (informational) ----------------------------
trial_offset_summary <- df %>%
  group_by(id, trial_number) %>%
  summarise(
    total_samples   = n(),
    offset_samples  = sum(aoi_label == "Offset", na.rm = TRUE),
    offset_prop     = mean(aoi_label == "Offset", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(offset_prop))

cat("Trials with the highest gaze-offset proportion (top 20):\n")
print(head(trial_offset_summary, 20))

# =============================================================================
# OUTPUT
# =============================================================================
# df_aoi  – clean eye-tracking data: trials with <= 50 % PD NAs AND gaze in AOI
# trial_na_summary    – per-trial NA proportions (all trials)
# trial_offset_summary – per-trial gaze-offset proportions (after NA exclusion)

cat("\n==========================================================\n")
cat("Quality checks complete.\n")
cat("  Clean dataset : 'df_aoi'\n")
cat(sprintf("  Rows          : %d\n", nrow(df_aoi)))
cat(sprintf("  Participants  : %d\n", length(unique(df_aoi$id))))
cat("==========================================================\n")

```

## Baseline PD

```{r echo=TRUE, fig.width=12, fig.height=8}
# Summary statistics of the baseline means by trial and participant 

summary_stats <- df_aoi %>%
  group_by(id,trial) %>%
  summarise(
    mean = mean(mean_baseline_pd, na.rm = TRUE),
    median = median(mean_baseline_pd, na.rm = TRUE),
    sd = sd(mean_baseline_pd, na.rm = TRUE),
    min = min(mean_baseline_pd, na.rm = TRUE),
    max = max(mean_baseline_pd, na.rm = TRUE),
    n = sum(!is.na(mean_baseline_pd))
  )
# Print a nice formatted table
summary_stats %>%
  kable("html", caption = "Summary Statistics of Mean Baseline PD by Participant and Trial") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1:2, bold = TRUE)

ggplot(df_aoi, aes(x = factor(id), y = mean_baseline_pd)) +
  geom_boxplot(outlier.color = "red", fill = "skyblue", alpha = 0.7) +
  facet_wrap(~trial, scales = "free_x", ncol = 3) +  # Adjust columns for better layout
  labs(
    title = "Baseline Pupil Diameter by Trial",
    subtitle = "Boxplot of mean baseline pupil diameter across participants",
    x = "Participant ID",
    y = "Mean Baseline Pupil Diameter"
  ) +
  theme_minimal(base_size = 14) +  # Larger base font size for readability
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 10, face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, margin = margin(b = 10)),
    strip.text = element_text(face = "bold", size = 12, color = "darkred"),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_blank()
  )

```

### exclude ID with low PD

```{r exclude_id}
## Current set up is to exclude specific ID
## for the future cutoff value (under 3)
ids_to_remove <- "LM_002"
df_aoi_clean <- df_aoi %>%
  filter(id != ids_to_remove)

df_trial_clean <- df_t_filtered %>%
  filter(id != ids_to_remove)

cat(
  "Exclude participants with a mean baseline pd < 3\n",
  "Before filtering:\n",
  "Number of unique participants: ", length(unique(df_aoi$id)), "\n",
  "Number of trials: ", sum(df_aoi$trial_number), "\n",
  "Number of rows in df_aoi: ", nrow(df_aoi), "\n\n",
  
  "After filtering:\n",
  "Number of unique participants: ", length(unique(df_aoi_clean$id)), "\n",
  "Number of trials: ", sum(df_aoi_clean$trial_number), "\n",
  "Number of rows in df_aoi_clean: ", nrow(df_aoi_clean), "\n\n",
  
  "trial data before filtering rows: ", nrow(df_t_filtered), "\n",
  "Trial data after filtering rows: ", nrow(df_trial_clean), "\n"
)

```
## pupillary response visualization

```{r , echo=FALSE, fig.width=10, fig.height=8}
p1 <-ggplot(df_aoi_clean, aes(x = ts_trial, y = pd_b_corr, fill= trial, color = trial)) +
  geom_smooth() +
  xlim(c(0,2))+
  labs(
    title = " baseline corrected PD Oddball vs. Standard, only AOI",
    x = "Time(s)",
    y = "Corrected PD"
  ) +
  theme_minimal()+
  theme(legend.position = "right")

p1 
```



