---
title: "LOCUS-MENTAL Eye Tracking Battery Validation"
author: "Iskra Todorova & Nico Bast"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi        = 150
)
```

---

# Setup

## Package Installation & Loading

```{r packages, }
pkgs <- c(
  "tidyverse", "GGally", "lme4", "emmeans", "lmerTest",
  "dplyr", "corrplot", "factoextra", "ggpubr", "Hmisc", "kableExtra", "broom"
)

installed_packages <- pkgs %in% rownames(installed.packages())
if (any(!installed_packages)) {
  install.packages(pkgs[!installed_packages])
}

lapply(pkgs, function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    message(paste("Package", pkg, "could not be loaded."))
  }
})
```

## File Paths

```{r paths}
home_path      <- "S:/KJP_Studien"
data_path_vo   <- "/LOCUS_MENTAL/6_Versuchsdaten/visual_oddball/"
data_path_ao   <- "/LOCUS_MENTAL/6_Versuchsdaten/auditory_oddball/"
data_path_rss  <- "/LOCUS_MENTAL/6_Versuchsdaten/rapid_sound_sequences/"
data_path_cvs  <- "/LOCUS_MENTAL/6_Versuchsdaten/visual_search_task/"
demo_data_path <- "S:/KJP_Studien/LOCUS_MENTAL/6_Versuchsdaten/"
```

---

# Data Loading

```{r load-data}
# Aggregated task data
df_ao  <- readRDS(paste0(home_path, data_path_ao,  "df_sepr_aggregated_AO.rds"))
df_vo  <- readRDS(paste0(home_path, data_path_vo,  "sepr_condition.rds"))
df_rss <- readRDS(paste0(home_path, data_path_rss, "df_sepr_aggregated_RSS.rds"))
df_cvs <- readRDS(paste0(home_path, data_path_cvs, "df_cued_agg.rds"))

# Baseline pupil size (BPS) trial-level data
bps_rss <- readRDS(paste0(home_path, data_path_rss, "bps_data.rds"))
bps_ao  <- readRDS(paste0(home_path, data_path_ao,  "bps_data.rds"))
bps_cvs <- readRDS(paste0(home_path, data_path_cvs, "bps_data.rds"))

# Demographic data
load(file.path(demo_data_path, "demo_data.rda"))
demo_data <- data
```

---

# Data Preparation

Each task dataset is cleaned and standardised before being joined into a single master dataframe.

## Auditory Oddball (AO)

```{r prep-ao}
data_ao <- df_ao %>%
  select(id, SEPR_AO_m_oddball, SEPR_AO_m_standard, BPS_oddball, BPS_standard) %>%
  rename(
    BPS_AO_standard = BPS_standard,
    BPS_AO_oddball  = BPS_oddball
  ) %>%
  mutate(
    SEPR_AO_standard_z = as.numeric(scale(SEPR_AO_m_standard)),
    SEPR_AO_oddball_z  = as.numeric(scale(SEPR_AO_m_oddball))
  )
```

## Rapid Sound Sequences (RSS)

```{r prep-rss}
data_rss <- df_rss %>%
  select(
    id,
    RSS_SEPR_early_mean_control, RSS_SEPR_early_mean_transition,
    RSS_BPS_mean_control, RSS_BPS_mean_transition,
    mean_diff_RAND_to_REG1, mean_diff_REG_to_RAND, mean_diff_RAND_to_REG10
  ) %>%
  rename(
    BPR_RSS_control    = RSS_BPS_mean_control,
    BPS_RSS_transition = RSS_BPS_mean_transition
  ) %>%
  mutate(
    RSS_transition_z = as.numeric(scale(RSS_SEPR_early_mean_transition)),
    RSS_control_z    = as.numeric(scale(RSS_SEPR_early_mean_control))
  )
```

## Visual Oddball (VO)

```{r prep-vo}
# Pivot from long to wide; retain only participants with >= 7 trials in both conditions
data_vo <- df_vo %>%
  filter(n_trials >= 7) %>%
  group_by(id) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  mutate(sepr_scaled = as.numeric(scale(sepr_mean))) %>%
  select(id, condition, sepr_mean, sepr_scaled, BPS, n_trials) %>%
  pivot_wider(
    names_from  = condition,
    values_from = c(sepr_mean, sepr_scaled, BPS, n_trials),
    names_glue  = "{.value}_VO_{condition}"
  )

cat("Participants remaining in VO task:", nrow(data_vo), "\n")
```

## Cued Visual Search (CVS)

```{r prep-cvs}
data_cvs <- df_cvs %>%
  select(
    id,
    CEPR_CUED_mean_cued, CEPR_CUED_mean_standard,
    SEPR_mean_cued, SEPR_mean_standard,
    BPS_cued, BPS_standard
  )
```

## Master Dataframe

```{r join-data}
# Full join retains all participants; missing tasks appear as NAs.
# VO excluded pending data structure harmonisation — see note above.
df_final <- data_ao %>%
  full_join(data_rss, by = "id") %>%
  # full_join(data_vo, by = "id") %>%
  full_join(data_cvs, by = "id")

cat("Total participants in master dataframe:", nrow(df_final), "\n")
```

> **Note:** The VO task is currently excluded from the master dataframe because the
> task effects did not worked out as expected probably due to task structure.
> However the BPS is included in the analysis.
> Re-enable the commented `full_join` line and remove the VO row from the BPS summary
> if the data is needed
---

# Baseline Pupil Size (BPS)

BPS reflects individual differences in tonic arousal and is expected to be highly stable
within a person across tasks. 

## Within-Battery Consistency

```{r bps-consistency}
bps_matrix <- df_final %>% select(id, starts_with("BPS"))

cor_bps <- cor(
  bps_matrix %>% select(-id),
  use    = "pairwise.complete.obs",
  method = "spearman"
)

cat("Spearman Correlations of Baseline Pupil Size Across Tasks:\n")
print(round(cor_bps, 2))
```

```{r bps-corrplot, fig.height=6, fig.width=7}
corrplot(
  cor_bps,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.srt      = 45,
  addCoef.col = "black",
  number.cex  = 0.8,
  col         = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  title       = "BPS Cross-Task Consistency (Spearman)",
  mar         = c(0, 0, 2, 0)
)
```

## BPS Descriptive Summary

```{r bps-summary}
bps_summary <- data.frame(
  Task     = c("AO", "RSS", "CVS"),
  Mean_BPS = c(
    mean(df_final$BPS_AO_standard,    na.rm = TRUE),
    mean(df_final$BPS_RSS_transition, na.rm = TRUE),
    mean(df_final$BPS_cued,           na.rm = TRUE)
  ),
  SD_BPS = c(
    sd(df_final$BPS_AO_standard,    na.rm = TRUE),
    sd(df_final$BPS_RSS_transition, na.rm = TRUE),
    sd(df_final$BPS_cued,           na.rm = TRUE)
  )
)

bps_summary %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Mean Baseline Pupil Size (mm) per Task") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Habituation Analysis

Mixed-effects models test whether BPS drifts systematically over the course of each task.
Expectation: negative estimates

## BPS Trajectory Plots

Trial-by-trial mean BPS for each task. A downward trend indicates habituation; a flat
profile indicates stable arousal across the session.

```{r bps-trajectories, fig.height=5, fig.width=12}
# AO
ao_traj <- bps_ao %>%
  group_by(trial_number) %>%
  summarise(mean_BPS = mean(BPS, na.rm = TRUE),
            se_BPS   = sd(BPS, na.rm = TRUE) / sqrt(n()),
            .groups  = "drop") %>%
  mutate(Task = "Auditory Oddball")

# CVS
cvs_traj <- bps_cvs %>%
  group_by(trial_number) %>%
  summarise(mean_BPS = mean(BPS, na.rm = TRUE),
            se_BPS   = sd(BPS, na.rm = TRUE) / sqrt(n()),
            .groups  = "drop") %>%
  mutate(Task = "Cued Visual Search")

# RSS
rss_traj <- bps_rss %>%
  rename(trial_number = Trial.Number) %>%
  group_by(trial_number) %>%
  summarise(mean_BPS = mean(BPS, na.rm = TRUE),
            se_BPS   = sd(BPS, na.rm = TRUE) / sqrt(n()),
            .groups  = "drop") %>%
  mutate(Task = "Rapid Sound Sequences")

bind_rows(ao_traj, cvs_traj, rss_traj) %>%
  ggplot(aes(x = trial_number, y = mean_BPS)) +
  geom_ribbon(aes(ymin = mean_BPS - se_BPS, ymax = mean_BPS + se_BPS),
              fill = "steelblue", alpha = 0.2) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_smooth(method = "lm", colour = "red", linetype = "dashed",
              linewidth = 0.7, se = FALSE) +
  facet_wrap(~ Task, scales = "free_x") +
  labs(
    title   = "Baseline Pupil Size Over Trials (Mean ± SE)",
    x       = "Trial Number",
    y       = "Mean BPS (mm)",
    caption = "Red dashed line = linear trend fit. Shaded band = ±1 SE across participants."
  ) +
  theme_bw(base_size = 12) +
  theme(strip.background = element_rect(fill = "grey90"))
```

## Auditory Oddball

```{r hab-ao}
bps_ao$condition <- as.factor(bps_ao$condition)

m_ao_hab <- lmer(BPS ~ trial_number + (1 | id), data = bps_ao)
cat("--- AO: Main effect of trial number ---\n")
print(anova(m_ao_hab))

m_ao_int <- lmer(BPS ~ trial_number * condition + (1 | id), data = bps_ao)
cat("--- AO: Trial number x Condition interaction ---\n")
print(anova(m_ao_int))
```

## Cued Visual Search

```{r hab-cvs}
bps_cvs$Condition    <- as.factor(bps_cvs$Condition)
bps_cvs$trial_number <- as.numeric(bps_cvs$trial_number)

m_cvs_hab <- lmer(BPS ~ trial_number + (1 | id), data = bps_cvs)
cat("--- CVS: Main effect of trial number ---\n")
print(anova(m_cvs_hab))

m_cvs_int <- lmer(BPS ~ trial_number * Condition + (1 | id), data = bps_cvs)
cat("--- CVS: Trial number x Condition interaction ---\n")
print(anova(m_cvs_int))
```

## Rapid Sound Sequences

```{r hab-rss}
bps_rss$Condition    <- as.factor(bps_rss$Condition)
bps_rss$Trial.Number <- as.numeric(bps_rss$Trial.Number)

m_rss_hab <- lmer(BPS ~ Trial.Number + (1 | id), data = bps_rss)
cat("--- RSS: Main effect of trial number ---\n")
print(anova(m_rss_hab))

m_rss_int <- lmer(BPS ~ Trial.Number * Condition + (1 | id), data = bps_rss)
cat("--- RSS: Trial number x Condition interaction ---\n")
print(anova(m_rss_int))
```

---

# Correlation Analysis

> **Method note:** Spearman rank correlations are used throughout as the primary measure,
> as pupil dilation data may deviate from normality.

## Build Correlation Dataset

```{r cor-data}
df_cor_rpd_raw <- df_final %>%
  select(
    id,
    SEPR_AO_m_oddball, SEPR_AO_m_standard,                                  # AO
    RSS_SEPR_early_mean_control, RSS_SEPR_early_mean_transition,             # RSS SEPR
    mean_diff_RAND_to_REG1, mean_diff_REG_to_RAND, mean_diff_RAND_to_REG10, # RSS diffs
    SEPR_mean_cued, SEPR_mean_standard,                                      # CVS SEPR
    CEPR_CUED_mean_cued, CEPR_CUED_mean_standard                             # CVS CEPR
  )

cat("N participants in correlation dataset:", nrow(df_cor_rpd_raw), "\n")
```

## Spearman Correlation Matrix with FDR-Corrected p-values

`Hmisc::rcorr()` computes r and p simultaneously. p-values are then corrected for
multiple comparisons using the Benjamini-Hochberg (FDR) procedure.

```{r spearman-fdr}
cor_obj <- Hmisc::rcorr(
  as.matrix(df_cor_rpd_raw %>% select(-id)),
  type = "spearman"
)

r_mat <- cor_obj$r
p_mat <- cor_obj$P

# Apply BH-FDR correction to all upper-triangle p-values
p_fdr_mat <- matrix(NA, nrow = nrow(p_mat), ncol = ncol(p_mat),
                    dimnames = dimnames(p_mat))
p_fdr_vec <- p.adjust(p_mat[upper.tri(p_mat)], method = "BH")
p_fdr_mat[upper.tri(p_fdr_mat)] <- p_fdr_vec
p_fdr_mat[lower.tri(p_fdr_mat)] <- t(p_fdr_mat)[lower.tri(p_fdr_mat)]
diag(p_fdr_mat)                 <- 1  # diagonal has no p-value; set to 1 so corrplot ignores it

cat("Spearman r matrix:\n")
print(round(r_mat, 2))
cat("\nFDR-corrected p-value matrix (BH):\n")
print(round(p_fdr_mat, 3))
```

```{r spearman-corrplot, fig.height=8, fig.width=9}
# 1. Get the number of variables to help with positioning
n <- ncol(r_mat)

# 2. Run corrplot WITHOUT addCoef.col
corrplot(
  r_mat,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.srt      = 45,
  tl.cex      = 0.8,
  addCoef.col = NULL,      # Change this to NULL to stop the overlap
  p.mat       = p_fdr_mat,
  sig.level   = 0.05,
  insig       = "pch",   
  pch         = 8,       
  pch.cex     = 1.2,      # Slightly smaller stars help readability
  pch.col     = "black",
  col         = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  title       = "Spearman Correlations (stars = FDR p < .05)",
  mar         = c(0, 0, 2, 0)
)

# 3. Manually add the coefficients with a slight downward offset
# This loops through the upper triangle (where column j > row i)
for (i in 1:(n-1)) {
  for (j in (i+1):n) {
    # x is the column (j), y is the row (reversed for corrplot: n - i + 1)
    # We use 'pos = 1' to place the text directly below the center point
    text(x = j, 
         y = n - i + 1, 
         labels = round(r_mat[i, j], 2), 
         pos = 1,       # pos=1 puts text BELOW the center; pos=3 puts it ABOVE
         offset = 0.6,  # adjust this number to move the text further up or down
         cex = 0.6, 
         col = "black")
  }
}
```

## Significant Correlation Pairs (FDR p < .05)

```{r sig-pairs}
vars <- colnames(r_mat)

data.frame(
  Var1  = vars[row(r_mat)[upper.tri(r_mat)]],
  Var2  = vars[col(r_mat)[upper.tri(r_mat)]],
  r     = round(r_mat[upper.tri(r_mat)], 3),
  p_fdr = round(p_fdr_mat[upper.tri(p_fdr_mat)], 4)
) %>%
  filter(p_fdr < 0.05) %>%
  arrange(p_fdr) %>%
  kable(caption = "Significant Spearman Correlations (FDR-corrected p < .05)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Pairwise Scatter Plot Matrix (GGally)

```{r ggpairs, fig.height=13, fig.width=13}
ggpairs(
  df_cor_rpd_raw %>% select(-id),
  columns = 1:(ncol(df_cor_rpd_raw) - 1),
  upper   = list(continuous = wrap("cor", method = "spearman", size = 3)),
  lower   = list(continuous = wrap("points", alpha = 0.3, size = 0.8)),
  diag    = list(continuous = wrap("densityDiag"))
) +
  theme_bw(base_size = 9) +
  labs(title = "Battery Validation: Spearman Correlations Across Pupil Tasks")
```

## Cross-Task Summary: Primary Condition Means

The most direct test of convergent validity: Spearman correlations between the single
primary outcome from each task (AO Oddball, RSS Transition, CVS Cued).

```{r summary-cors}
cross_task_cor <- df_final %>%
  select(
    AO_Oddball     = SEPR_AO_m_oddball,
    RSS_Transition = RSS_SEPR_early_mean_transition,
    Cued_Task      = CEPR_CUED_mean_cued
  ) %>%
  as.matrix() %>%
  Hmisc::rcorr(type = "spearman")

cat("Cross-task r:\n")
print(round(cross_task_cor$r, 2))
cat("\nCross-task p (uncorrected — only 3 pairs):\n")
print(round(cross_task_cor$P, 3))
```

---

# Principal Component Analysis (PCA)

PCA summarises the shared variance structure across all task conditions and helps
identify whether tasks cluster into interpretable groups or measure largely independent
processes.

## Run PCA

```{r pca-run}
pca_data       <- df_cor_rpd_raw[, 2:ncol(df_cor_rpd_raw)]
pca_data_clean <- na.omit(pca_data)

cat("N entering PCA (complete cases):", nrow(pca_data_clean),
    "of", nrow(pca_data), "total\n")

pca_result <- prcomp(pca_data_clean, center = TRUE, scale. = TRUE)
```

## Scree Plot

```{r pca-scree}
fviz_eig(pca_result, addlabels = TRUE, ncp = 6) +
  geom_hline(yintercept = 100 / ncol(pca_data_clean),
             linetype = "dashed", colour = "grey50") +
  labs(
    title   = "Scree Plot: Variance Explained by Each Component",
    caption = "Dashed line = variance expected per variable under a null model (equal share)"
  ) +
  theme_bw()
```

## Variable Factor Map

```{r pca-var}
fviz_pca_var(
  pca_result,
  col.var       = "contrib",
  gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
  repel         = TRUE
) +
  labs(title = "Task Groupings (PCA Variable Factor Map)") +
  theme_bw()
```

## Component Loadings (PC1-PC3)

```{r pca-loadings}
loadings_df <- as.data.frame(round(pca_result$rotation[, 1:3], 3))

loadings_df %>%
  kable(caption = "PCA Loadings - First Three Components") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(abs(loadings_df$PC1) > 0.3), bold = TRUE)
```

---

# Demographic & Clinical Associations

## Merge Demographic Data

```{r demo-merge}
demo_vars <- demo_data %>%
  rename(id = ID) %>%
  select(id, sex, CBCL_T_INT, CBCL_T_EXT, CBCL_T_GES, IQ_verbal_z, IQ_nonverbal_z)

bps_matrix_d     <- bps_matrix     %>% inner_join(demo_vars, by = "id")
df_cor_rpd_raw_d <- df_cor_rpd_raw %>% inner_join(demo_vars, by = "id")

cat("N with complete demographic data:", nrow(bps_matrix_d), "\n")
```

## Correlations: BPS x Demographics

```{r cor-bps-demo, fig.height=7, fig.width=8}
# Setup data
cor_data      <- bps_matrix_d %>%
  mutate(sex = as.numeric(as.factor(sex))) %>%
  select(-id)

bps_demo_cor  <- Hmisc::rcorr(as.matrix(cor_data), type = "spearman")
bps_p <- bps_demo_cor$P
diag(bps_p) <- 1 

# 1. Plot without automatic coefficients
corrplot(
  bps_demo_cor$r,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.srt      = 45,
  tl.cex      = 0.8,
  addCoef.col = NULL,      # Disabled to prevent overlap
  p.mat       = bps_p,
  sig.level   = 0.05,
  insig       = "label_sig",
  col         = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  title       = "BPS x Demographics (Spearman, stars = p < .05)",
  mar         = c(0, 0, 2, 0)
)

# 2. Add coefficients manually with an offset
n_bps <- ncol(bps_demo_cor$r)
for (i in 1:(n_bps-1)) {
  for (j in (i+1):n_bps) {
    text(x = j, 
         y = n_bps - i + 1, 
         labels = sprintf("%.2f", bps_demo_cor$r[i, j]), 
         pos = 1,      # Places text BELOW the center/star
         offset = 0.7, # Adjust this if you want the number higher or lower
         cex = 0.65, 
         col = "black")
  }
}
```

## Correlations: SEPR x Demographics

```{r cor-sepr-demo, fig.height=8, fig.width=9}
# Setup data
cor_df        <- df_cor_rpd_raw_d %>%
  mutate(sex = as.numeric(as.factor(sex))) %>%
  select(-id)

sepr_demo_cor <- Hmisc::rcorr(as.matrix(cor_df), type = "spearman")
sepr_p <- sepr_demo_cor$P
diag(sepr_p) <- 1 

# 1. Plot without automatic coefficients
corrplot(
  sepr_demo_cor$r,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.srt      = 45,
  tl.cex      = 0.75,
  addCoef.col = NULL,      # Disabled to prevent overlap
  p.mat       = sepr_p,
  sig.level   = 0.05,
  insig       = "label_sig",
  col         = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  title       = "SEPR x Demographics (Spearman, stars = p < .05)",
  mar         = c(0, 0, 2, 0)
)

# 2. Add coefficients manually with an offset
n_sepr <- ncol(sepr_demo_cor$r)
for (i in 1:(n_sepr-1)) {
  for (j in (i+1):n_sepr) {
    text(x = j, 
         y = n_sepr - i + 1, 
         labels = sprintf("%.2f", sepr_demo_cor$r[i, j]), 
         pos = 1,      # Places text BELOW the center/star
         offset = 0.7, # Adjust as needed
         cex = 0.6, 
         col = "black")
  }
}
```

---

# Session Info

```{r session-info}
sessionInfo()
```
