---
title: "Sample Description"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
   
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,   # hide code in output — set to TRUE if you want to show it
  warning = FALSE,
  message = FALSE
)
options(knitr.table.format = "html")
```


```{r packages, include=FALSE}
pkgs <- c("readxl", "data.table", "ggplot2", "dplyr", "stringr",
          "psych", "lubridate", "tidyr", "knitr", "kableExtra", "ggcorrplot", 'ggVennDiagram')

installed_packages <- pkgs %in% rownames(installed.packages())
if (any(!installed_packages)) {
  install.packages(pkgs[!installed_packages])
}

lapply(pkgs, function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE))
    message(paste("Package", pkg, "not found."))
})
```


```{r load-data, include=FALSE}
data_path <- "//192.168.88.212/daten/KJP_Studien/LOCUS_MENTAL/6_Versuchsdaten/"
load(file.path(data_path, "demo_data.rda"))

# Convert age to months
data <- data %>%
  mutate(age_months = round(age * 12, 1))
```

# Sample Description Table
```{r compute-all, include=FALSE}
#sex

data$sex <- factor(data$sex,
                   levels = c("männlich", "weiblich"),
                   labels = c("male", "female"))

sex_counts <- table(data$sex)

# Chi-square goodness-of-fit (vs. 50/50)
chi_sex <- chisq.test(sex_counts)


sex_row <- data.frame(
  Variable = "Sex (female / male)",
  N        = as.character(sum(!is.na(data$sex))),
  M        = paste0(sex_counts[["female"]], " / ", sex_counts[["male"]]),
  SD       = paste0("(", round(sex_counts[["female"]] / sum(sex_counts) * 100, 1),
                    "% / ",
                    round(sex_counts[["male"]]   / sum(sex_counts) * 100, 1), "%)"),
  Min      = "—",
  Max      = "—",
  `SW W`   = paste0("\u03c7\u00b2(1) = ", round(chi_sex$statistic, 2)),
  `SW p`   = as.character(round(chi_sex$p.value, 3)),
  check.names = FALSE
) %>%
  mutate(across(everything(), as.character))


# continuous varibales
cont_vars <- c(
  "age_months", "IQ_verbal", "IQ_nonverbal",
  "CBCL_T_INT", "CBCL_T_EXT", "CBCL_T_GES",
  "ARI_Eltern_Score_total", "BIQ_Z_Score",
  "CBQ_Offenheit_Summenwert", "CBQ_Negativer_Affekt_Summenwert",
  "CBQ_Kontrollfaehigkeit_Summenwert", "CBQ_Ges_Summenwert",
  "F_SozU_Score", "ACE_Score_total"
)

desc_table <- data %>%
  select(all_of(cont_vars)) %>%
  summarise(across(
    everything(),
    list(
      N   = ~ sum(!is.na(.)),
      M   = ~ round(mean(., na.rm = TRUE), 2),
      SD  = ~ round(sd(.,   na.rm = TRUE), 2),
      Min = ~ round(min(.,  na.rm = TRUE), 2),
      Max = ~ round(max(.,  na.rm = TRUE), 2)
    ),
    .names = "{.col}__{.fn}"
  )) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "__") %>%
  pivot_wider(names_from = Stat, values_from = value)


# 3. Normality (Shapiro-Wilk)

sw_results <- data %>%
  select(all_of(cont_vars)) %>%
  summarise(across(
    everything(),
    list(
      W = ~ tryCatch(round(shapiro.test(na.omit(.))$statistic, 3),
                     error = function(e) NA_real_),
      p = ~ tryCatch(round(shapiro.test(na.omit(.))$p.value,   3),
                     error = function(e) NA_real_)
    ),
    .names = "{.col}__{.fn}"
  )) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "__") %>%
  pivot_wider(names_from = Stat, values_from = value) %>%
  mutate(Normal = case_when(
    is.na(p) ~ "-",
    p >= .05 ~ "yes",
    TRUE     ~ "no"
  ))

sw_results <- sw_results %>%
  mutate(p = ifelse(p == 0, "< .001", as.character(p)))


# Final Table

var_labels <- c(
  "age_months"                        = "Age (months)",
  "IQ_verbal"                         = "IQ Verbal",
  "IQ_nonverbal"                      = "IQ Nonverbal",
  "CBCL_T_INT"                        = "CBCL Internalizing (T)",
  "CBCL_T_EXT"                        = "CBCL Externalizing (T)",
  "CBCL_T_GES"                        = "CBCL Total (T)",
  "ARI_Eltern_Score_total"            = "ARI Parent Total",
  "BIQ_Z_Score"                       = "BIQ Z-Score",
  "CBQ_Offenheit_Summenwert"          = "CBQ Openness",
  "CBQ_Negativer_Affekt_Summenwert"   = "CBQ Negative Affect",
  "CBQ_Kontrollfaehigkeit_Summenwert" = "CBQ Effortful Control",
  "CBQ_Ges_Summenwert"                = "CBQ Total",
  "F_SozU_Score"                      = "F-SozU Social Support",
  "ACE_Score_total"                   = "ACE Total Score"
)

final_table <- desc_table %>%
  left_join(sw_results %>% select(Variable, W, p), by = "Variable") %>%
  mutate(Variable = recode(Variable, !!!var_labels)) %>%
  rename("SW W" = W, "SW p" = p) %>%
  mutate(across(everything(), as.character))

final_table_full <- bind_rows(sex_row, final_table)
```

```{r final-table}
kable(final_table_full,
      align   = "lrrrrrrr",
      caption = "Table 1. Sample Description") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width        = FALSE
  ) %>%
  pack_rows("Demographics & Cognition", 1, 4) %>%
  pack_rows("CBCL",                     5, 7) %>%
  pack_rows("Questionnaires",           8, 15) %>%
  footnote(general = paste0(
    "For Sex: M = N female / N male; SD = (% female / % male); ",
    "SW W = \u03c7\u00b2(1) goodness-of-fit test against equal distribution. ",
    "For continuous variables: SW W and SW p refer to the Shapiro-Wilk normality test."
  ))
```

# Figures

```{r plot-theme, include=FALSE}
# Shared clean theme for all plots
theme_desc <- theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold", size = 12),
    axis.title   = element_text(size = 10),
    panel.grid.minor = element_blank()
  )
```

## Sex Distribution

```{r plot-sex, fig.width=4, fig.height=4}
data %>%
  count(sex) %>%
  mutate(pct = round(n / sum(n) * 100, 1),
         label = paste0(n, "\n(", pct, "%)")) %>%
  ggplot(aes(x = sex, y = n, fill = sex)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("female" = "#5B8DB8", "male" = "#E08B5D")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Figure 1. Sex Distribution",
       x = NULL, y = "Count") +
  theme_desc
```

## Age Distribution

```{r plot-age, fig.width=5, fig.height=4}
ggplot(data, aes(x = age_months)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 6, fill = "#5B8DB8", color = "white", alpha = 0.8) +
  stat_function(fun  = dnorm,
                args = list(mean = mean(data$age_months, na.rm = TRUE),
                            sd   = sd(data$age_months,   na.rm = TRUE)),
                color = "#E08B5D", linewidth = 1) +
  labs(title = "Figure 2. Age Distribution",
       x = "Age (months)", y = "Density") +
  theme_desc
```

## IQ Scores

```{r plot-iq, fig.width=5, fig.height=4}
data %>%
  select(IQ_verbal, IQ_nonverbal) %>%
  pivot_longer(everything(),
               names_to  = "Scale",
               values_to = "Score") %>%
  mutate(Scale = recode(Scale,
    "IQ_verbal"    = "Verbal",
    "IQ_nonverbal" = "Nonverbal")) %>%
  ggplot(aes(x = Score, fill = Scale, color = Scale)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  scale_fill_manual(values  = c("Verbal" = "#5B8DB8", "Nonverbal" = "#E08B5D")) +
  scale_color_manual(values = c("Verbal" = "#5B8DB8", "Nonverbal" = "#E08B5D")) +
  labs(title = "Figure 3. IQ Score Distributions",
       x = "Raw Score", y = "Density", fill = NULL, color = NULL) +
  theme_desc +
  theme(legend.position = "bottom")
```

## CBCL Clinical Symptoms (T ≥ 65)

```{r plot-cbcl, fig.width=6, fig.height=4}
data %>%
  select(CBCL_T_INT, CBCL_T_EXT, CBCL_T_GES) %>%
  pivot_longer(everything(),
               names_to  = "Scale",
               values_to = "T_Score") %>%
  mutate(Scale = recode(Scale,
    "CBCL_T_INT" = "Internalizing",
    "CBCL_T_EXT" = "Externalizing",
    "CBCL_T_GES" = "Total")) %>%
  ggplot(aes(x = Scale, y = T_Score, fill = Scale)) +
  
  #show as violinplot with individual points jittered
  geom_violin(alpha = 0.7, outlier.shape = 21,
               outlier.fill = "white", show.legend = FALSE) +
  geom_jitter(aes(color=Scale)) +
  
  # clinical cutoff T ≥ 60
  geom_hline(yintercept = 65, linetype = "dashed",
             color = "grey40", linewidth = 0.5) +
  
  annotate("text", x = 0.55, y = 61.5,
           label = "T = 65", size = 3, color = "grey40") +
  scale_fill_manual(values = c(
    "Internalizing" = "#5B8DB8",
    "Externalizing" = "#E08B5D",
    "Total"         = "#6DBF8B")) +
  scale_color_manual(values = c(
    "Internalizing" = "#5B8DB8",
    "Externalizing" = "#E08B5D",
    "Total"         = "#6DBF8B")) +
  
  labs(title = "Figure 4. CBCL T-Score Distributions",
       x = NULL, y = "T-Score") +
  theme_desc
```


```{r cbcl-combined, fig.width=6, fig.height=5}

cutoff <- 65

# ── Flag participants ──────────────────────────────────────────────────────────
cbcl_flags <- data %>%
  select(ID, CBCL_T_INT, CBCL_T_EXT, CBCL_T_GES) %>%
  filter(!is.na(CBCL_T_INT) & !is.na(CBCL_T_EXT)) %>%
  mutate(
    clinical_INT = CBCL_T_INT >= cutoff,
    clinical_EXT = CBCL_T_EXT >= cutoff,
    clinical_GES = CBCL_T_GES >= cutoff
  )

# ── Per-scale summary ──────────────────────────────────────────────────────────
scale_summary <- cbcl_flags %>%
  summarise(
    across(c(clinical_INT, clinical_EXT, clinical_GES),
           list(n   = ~ sum(.),
                pct = ~ round(mean(.) * 100, 1)))
  ) %>%
  pivot_longer(everything(),
               names_to  = c("Scale", "Stat"),
               names_sep = "_(?=[^_]+$)") %>%   # split on last underscore
  pivot_wider(names_from = Stat, values_from = value) %>%
  mutate(
    Scale = recode(Scale,
      "clinical_INT" = "Internalizing (T ≥ 65)",
      "clinical_EXT" = "Externalizing (T ≥ 65)",
      "clinical_GES" = "Total Problems (T ≥ 65)"),
    N_total = nrow(cbcl_flags)
  ) %>%
  select(Group = Scale, `N clinical` = n, `N total` = N_total, `%` = pct)

# ── Mutually exclusive groups ──────────────────────────────────────────────────
overlap_summary <- cbcl_flags %>%
  mutate(Group = case_when(
    clinical_INT &  clinical_EXT ~ "Both INT & EXT",
    clinical_INT & !clinical_EXT ~ "INT only",
   !clinical_INT &  clinical_EXT ~ "EXT only",
    TRUE                          ~ "Neither"
  )) %>%
  count(Group) %>%
  mutate(
    `N total`    = nrow(cbcl_flags),
    `%`          = round(n / nrow(cbcl_flags) * 100, 1),
    `N clinical` = n
  ) %>%
  arrange(match(Group, c("Both INT & EXT", "INT only", "EXT only", "Neither"))) %>%
  select(Group, `N clinical`, `N total`, `%`)
```

```{r combined-table}
# ── One combined table ─────────────────────────────────────────────────────────
combined_table <- bind_rows(scale_summary, overlap_summary)

kable(combined_table,
      align   = "lrrr",
      caption = "Table 2. CBCL Clinically Relevant Symptoms (T ≥ 65)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  pack_rows("By Scale",            1, 3) %>%
  pack_rows("Symptom Profile",     4, 7) %>%
  footnote(general = paste0(
    "T ≥ 65 = borderline clinical cutoff. ",
    "Symptom Profile shows mutually exclusive groups based on ",
    "Internalizing (INT) and Externalizing (EXT) scores. ",
    "Total Problems is reported separately as it is a composite scale."
  ))
```

```{r figure5, fig.width=6, fig.height=5}
# ── Figure 5: Stacked bar — clinical vs subclinical ───────────────────────────
scale_summary %>%
  mutate(pct_sub = 100 - `%`,
         Group   = factor(Group, levels = c(
           "Internalizing (T ≥ 65)",
           "Externalizing (T ≥ 65)",
           "Total Problems (T ≥ 65)"))) %>%
  pivot_longer(cols      = c(`%`, pct_sub),
               names_to  = "Clinical",
               values_to = "Percent") %>%
  mutate(Clinical = recode(Clinical,
    "%"       = "Clinical (T \u2265 65)",
    "pct_sub" = "Subclinical (T < 65)"),
    Clinical = factor(Clinical,
      levels = c("Subclinical (T < 65)", "Clinical (T \u2265 65)"))) %>%
  ggplot(aes(x = Group, y = Percent, fill = Clinical)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = paste0(Percent, "%")),
            position = position_stack(vjust = 0.5),
            size = 3.5, color = "white", fontface = "bold") +
  scale_fill_manual(values = c(
    "Subclinical (T < 65)"   = "#A8C8E8",
    "Clinical (T \u2265 65)" = "#C0392B")) +
  scale_x_discrete(labels = c(
    "Internalizing (T ≥ 65)" = "Internalizing",
    "Externalizing (T ≥ 65)" = "Externalizing",
    "Total Problems (T ≥ 65)"= "Total")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title    = "Figure 5. CBCL Clinically Relevant Symptoms (T \u2265 65)",
       subtitle = paste0("N = ", nrow(cbcl_flags)),
       x = NULL, y = "Percentage of Participants", fill = NULL) +
  theme_desc +
  theme(legend.position = "bottom")
```

```{r figure6, fig.width=6, fig.height=4.5}
# ── Figure 6: Mutually exclusive symptom profiles ─────────────────────────────
overlap_summary %>%
  mutate(Group = factor(Group,
    levels = c("Both INT & EXT", "INT only", "EXT only", "Neither"))) %>%
  ggplot(aes(x = Group, y = `%`, fill = Group)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = paste0(`N clinical`, "\n(", `%`, "%)")),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c(
    "Both INT & EXT" = "#C0392B",
    "INT only"       = "#E08B5D",
    "EXT only"       = "#5B8DB8",
    "Neither"        = "#A8C8E8")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.2))) +
  labs(title    = "Figure 6. Mutually Exclusive CBCL Symptom Profiles (T \u2265 65)",
       subtitle = paste0("N = ", nrow(cbcl_flags)),
       x = NULL, y = "% of Participants") +
  theme_desc
```

## Additional Variable Distributions

```{r plot-ari, fig.width=5, fig.height=4}
# ── ARI: cutoff ≥ 4 ───────────────────────────────────────────────────────────
ari_cutoff  <- 4
n_ari_valid <- sum(!is.na(data$ARI_Eltern_Score_total))
n_ari_atrisk <- sum(data$ARI_Eltern_Score_total >= ari_cutoff, na.rm = TRUE)  
pct_ari_atrisk <- round(n_ari_atrisk / n_ari_valid * 100, 1)

data %>%
  filter(!is.na(ARI_Eltern_Score_total)) %>%
  count(ARI_Eltern_Score_total) %>%
  mutate(elevated = ARI_Eltern_Score_total >= ari_cutoff) %>%  # >= here
  ggplot(aes(x = factor(ARI_Eltern_Score_total), y = n, fill = elevated)) +
  geom_col(color = "white", alpha = 0.85, width = 0.7, show.legend = TRUE) +
  geom_vline(xintercept = ari_cutoff + 0.5,         # line now sits before 4
             linetype = "dashed", color = "#C0392B", linewidth = 0.8) +
  annotate("text", x = ari_cutoff + 0.2, y = Inf,
           label = paste0("At-risk\n(n = ", n_ari_atrisk, ")"),
           vjust = 1.5, hjust = 0, size = 3, color = "#C0392B") +
  scale_fill_manual(
    values = c("FALSE" = "#5B8DB8", "TRUE" = "#C0392B"),
    labels = c("FALSE" = paste0("Not elevated (< ",   ari_cutoff, ")"),
               "TRUE"  = paste0("Elevated (\u2265 ",  ari_cutoff, ")"))  # >= here
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "Figure 8. ARI Parent Total Score",
    subtitle = paste0("N = ", n_ari_valid,
                      " | At-risk (\u2265 ", ari_cutoff, "): n = ", n_ari_atrisk,
                      " (", pct_ari_atrisk, "%)"),
    x    = "ARI Total Score",
    y    = "Count",
    fill = NULL
  ) +
  theme_desc +
  theme(legend.position = "bottom")
```

```{r plot-biq, fig.width=5, fig.height=4}
# ── BIQ: cutoff ≥ 2 ───────────────────────────────────────────────────────────
biq_cutoff   <- 2
n_biq_valid  <- sum(!is.na(data$BIQ_Z_Score))
n_biq_atrisk <- sum(data$BIQ_Z_Score >= biq_cutoff, na.rm = TRUE)
pct_biq_atrisk <- round(n_biq_atrisk / n_biq_valid * 100, 1)

ggplot(data, aes(x = BIQ_Z_Score)) +
  geom_histogram(aes(y = after_stat(density), fill = after_stat(x) >= biq_cutoff),
                 binwidth = 0.3, color = "white", alpha = 0.85) +
  stat_function(fun  = dnorm,
                args = list(mean = mean(data$BIQ_Z_Score, na.rm = TRUE),
                            sd   = sd(data$BIQ_Z_Score,   na.rm = TRUE)),
                color = "#E08B5D", linewidth = 1) +
  geom_vline(xintercept = biq_cutoff,
             linetype = "dashed", color = "#C0392B", linewidth = 0.8) +
  annotate("text", x = biq_cutoff + 0.05, y = Inf,
           label = paste0("At-risk\n(n = ", n_biq_atrisk, ")"),
           vjust = 1.5, hjust = 0, size = 3, color = "#C0392B") +
  scale_fill_manual(
    values = c("FALSE" = "#5B8DB8", "TRUE" = "#C0392B"),
    labels = c("FALSE" = paste0("Not elevated (< ",  biq_cutoff, ")"),
               "TRUE"  = paste0("Elevated (\u2265 ", biq_cutoff, ")"))
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "Figure 9. BIQ Z-Score Distribution",
    subtitle = paste0("N = ", n_biq_valid,
                      " | At-risk (\u2265 ", biq_cutoff, "): n = ", n_biq_atrisk,
                      " (", pct_biq_atrisk, "%)"),
    x    = "BIQ Z-Score",
    y    = "Density",
    fill = NULL
  ) +
  theme_desc +
  theme(legend.position = "bottom")
```

```{r plot-cbq, fig.width=8, fig.height=6}
# ── CBQ subscales — faceted ────────────────────────────────────────────────────
data %>%
  select(CBQ_Offenheit_Summenwert, CBQ_Negativer_Affekt_Summenwert,
         CBQ_Kontrollfaehigkeit_Summenwert, CBQ_Ges_Summenwert) %>%
  pivot_longer(everything(), names_to = "Scale", values_to = "Score") %>%
  mutate(Scale = recode(Scale,
    "CBQ_Offenheit_Summenwert"          = "Openness",
    "CBQ_Negativer_Affekt_Summenwert"   = "Negative Affect",
    "CBQ_Kontrollfaehigkeit_Summenwert" = "Effortful Control",
    "CBQ_Ges_Summenwert"                = "Total")) %>%
  group_by(Scale) %>%
  mutate(m  = mean(Score, na.rm = TRUE),
         sd = sd(Score,   na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = Score)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 5, fill = "#5B8DB8",
                 color = "white", alpha = 0.85) +
  stat_function(
    fun  = dnorm,
    aes(color = "Normal curve"),
    args = function(x) list(
      mean = mean(x$Score, na.rm = TRUE),
      sd   = sd(x$Score,   na.rm = TRUE)
    ),
    linewidth = 0.8
  ) +
  # Add M ± SD as subtitle per facet via geom_text
  geom_text(
    data = . %>%
      group_by(Scale) %>%
      summarise(m = mean(Score, na.rm=TRUE),
                sd = sd(Score, na.rm=TRUE),
                .groups = "drop") %>%
      mutate(label = paste0("M = ", round(m,1), ", SD = ", round(sd,1)),
             x = -Inf, y = Inf),
    aes(x = x, y = y, label = label),
    hjust = -0.1, vjust = 1.5, size = 3, inherit.aes = FALSE
  ) +
  facet_wrap(~ Scale, scales = "free_x", ncol = 2) +
  scale_color_manual(values = c("Normal curve" = "#E08B5D")) +
  labs(title = "Figure 9. CBQ Subscale Distributions",
       x = "Sum Score", y = "Density", color = NULL) +
  theme_desc +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

```{r plot-fsozu, fig.width=5, fig.height=4}
# ── F-SozU ─────────────────────────────────────────────────────────────────────
ggplot(data, aes(x = F_SozU_Score)) +
  geom_histogram(binwidth = 2, fill = "#5B8DB8",
                 color = "white", alpha = 0.85) +
  labs(title    = "Figure 10. F-SozU Social Support Score",
       subtitle = paste0("M = ", round(mean(data$F_SozU_Score, na.rm=TRUE), 2),
                         ", SD = ", round(sd(data$F_SozU_Score, na.rm=TRUE), 2)),
       x = "F-SozU Score", y = "Count") +
  theme_desc
```

```{r plot-ace, fig.width=5, fig.height=4}
# ── ACE — bar chart (discrete count 0–10) ─────────────────────────────────────
data %>%
  filter(!is.na(ACE_Score_total)) %>%
  count(ACE_Score_total) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  ggplot(aes(x = factor(ACE_Score_total), y = n)) +
  geom_col(fill = "#5B8DB8", color = "white", alpha = 0.85, width = 0.7) +
  geom_text(aes(label = paste0(n, "\n(", pct, "%)")),
            vjust = -0.3, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(title    = "Figure 11. ACE Total Score Distribution",
       subtitle = paste0("N = ", sum(!is.na(data$ACE_Score_total)),
                         " | M = ", round(mean(data$ACE_Score_total, na.rm=TRUE), 2),
                         ", SD = ", round(sd(data$ACE_Score_total, na.rm=TRUE), 2)),
       x = "Number of Adverse Childhood Experiences", y = "Count") +
  theme_desc
```

```{r ace-items, fig.width=8, fig.height=6}
# ── ACE variables ──────────────────────────────────────────────────────────────
ace_vars <- c(
  "emotional_abuse", "physical_abuse", "sexual_assult",
  "emotional_neglect", "physical_neglect", "parents_separation",
  "mother_treated_violently", "family_drug_alc_problem",
  "family_mental_illness", "parent_prison", "low_ses",
  "high_peer_victimization", "high_peer_social_isolation",
  "high_exposute_community_violence"
)

# ── Count responses ────────────────────────────────────────────────────────────
ace_summary <- data %>%
  select(all_of(ace_vars)) %>%
  summarise(across(everything(),
    list(yes   = ~ sum(. == 1, na.rm = TRUE),
         total = ~ sum(!is.na(.))),
    .names = "{.col}__{.fn}")) %>%
  pivot_longer(everything(),
               names_to  = c("Item", "Stat"),
               names_sep = "__") %>%
  pivot_wider(names_from = Stat, values_from = value) %>%
  mutate(
    pct  = round(yes / total * 100, 1),
    Item = str_to_title(str_replace_all(Item, "_", " "))
  ) %>%
  arrange(desc(pct))

# ── Plot ───────────────────────────────────────────────────────────────────────
ggplot(ace_summary, aes(x = reorder(Item, pct), y = pct, fill = pct)) +
  geom_col(color = "white", alpha = 0.85) +
  geom_text(aes(label = paste0(yes, "/", total, "\n(", pct, "%)")),
            hjust = -0.05, size = 3, fontface = "bold") +
  scale_fill_gradient(low = "#A8C8E8", high = "#C0392B",
                      name = "Prevalence (%)") +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.18))) +
  coord_flip() +
  labs(
    title    = "Figure 18. Prevalence of Adverse Childhood Experiences",
    subtitle = paste0("N = ", max(ace_summary$total),
                      " | % of sample reporting 'yes' (1) for each ACE"),
    x        = NULL,
    y        = "Percentage (%)"
  ) +
  theme_desc +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 9)
  )
```

## Correlation Matrix
```{r analysis-vars, include=FALSE}
# ── Variable selection ─────────────────────────────────────────────────────────
# Note: CBCL_T_GES excluded from regression (composite of INT + EXT → multicollinearity)
# but kept in correlation matrix for completeness

all_vars <- c(
  "age_months", "IQ_verbal", "IQ_nonverbal",
  "CBCL_T_INT", "CBCL_T_EXT", "CBCL_T_GES",
  "ARI_Eltern_Score_total", "BIQ_Z_Score",
  "CBQ_Offenheit_Summenwert", "CBQ_Negativer_Affekt_Summenwert",
  "CBQ_Kontrollfaehigkeit_Summenwert", "CBQ_Ges_Summenwert",
  "F_SozU_Score", "ACE_Score_total"
)

# Clean labels for plots
var_labels_short <- c(
  "age_months"                        = "Age",
  "IQ_verbal"                         = "IQ verbal",
  "IQ_nonverbal"                      = "IQ nonverbal",
  "CBCL_T_INT"                        = "INT",
  "CBCL_T_EXT"                        = "EXT",
  "CBCL_T_GES"                        = "Total",
  "ARI_Eltern_Score_total"            = "ARI",
  "BIQ_Z_Score"                       = "BIQ",
  "CBQ_Offenheit_Summenwert"          = "CBQ Open.",
  "CBQ_Negativer_Affekt_Summenwert"   = "CBQ Neg. Aff.",
  "CBQ_Kontrollfaehigkeit_Summenwert" = "CBQ Eff. Ctrl.",
  "CBQ_Ges_Summenwert"                = "CBQ Total",
  "F_SozU_Score"                      = "F-SozU",
  "ACE_Score_total"                   = "ACE"
)

# Keep original names — used for regression
analysis_data <- data %>% select(all_of(all_vars))

# Renamed version — used only for correlation matrix display
analysis_data_display <- analysis_data
names(analysis_data_display) <- var_labels_short[names(analysis_data_display)]
```
```{r corr-matrix, fig.width=9, fig.height=8}
# ── Spearman correlations (conservative choice given mixed normality) ───────────
corr_results <- psych::corr.test(
  analysis_data_display,
  method = "spearman",
  adjust = "BH"       # Benjamini-Hochberg correction for multiple comparisons
)

r_matrix <- corr_results$r
p_raw    <- corr_results$p

# Extract lower triangle, adjust, then mirror to upper triangle
p_adjusted <- matrix(NA, nrow = nrow(p_raw), ncol = ncol(p_raw),
                     dimnames = dimnames(p_raw))

p_vals_adj <- p.adjust(p_raw[lower.tri(p_raw)], method = "BH")
p_adjusted[lower.tri(p_adjusted)] <- p_vals_adj
p_adjusted[upper.tri(p_adjusted)] <- t(p_adjusted)[upper.tri(p_adjusted)]
diag(p_adjusted) <- 0    # diagonal gets 0 so ggcorrplot handles it cleanly

ggcorrplot(
  r_matrix,
  type          = "upper",
  lab           = TRUE,
  lab_size      = 3,
  p.mat         = p_adjusted,
  sig.level     = 0.05,
  insig         = "blank",
  colors        = c("#C0392B", "white", "#2E86AB"),
  outline.color = "white",
  tl.cex        = 10,
  tl.srt        = 45
) +
  labs(
    title    = "Figure 15. Spearman Correlation Matrix",
    subtitle = "Only significant correlations shown (BH-corrected, p < .05)",
    fill     = "r"
  ) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "grey40")
  )
```
```

```{r corr-table}
# ── Printable correlation table ────────────────────────────────────────────────
corr_tidy <- as.data.frame(r_matrix) %>%
  tibble::rownames_to_column("Variable") %>%
  pivot_longer(-Variable, names_to = "Variable2", values_to = "r") %>%
  left_join(
    as.data.frame(p_adjusted) %>%
      tibble::rownames_to_column("Variable") %>%
      pivot_longer(-Variable, names_to = "Variable2", values_to = "p_adj"),
    by = c("Variable", "Variable2")
  ) %>%
  filter(Variable != Variable2) %>%
  # Keep only lower triangle
  filter(match(Variable,  names(var_labels_short)) >
         match(Variable2, names(var_labels_short))) %>%
  mutate(
    r     = round(r, 3),
    p_adj = ifelse(p_adj < .001, "< .001", round(p_adj, 3)),
    sig   = ifelse(p_adj == "< .001" | as.numeric(p_adj) < .05, "*", "")
  ) %>%
  arrange(Variable, Variable2)

knitr::kable(corr_tidy,
      col.names = c("Variable 1", "Variable 2", "r", "p (adj.)", ""),
      caption   = "Table 3. Spearman Correlations (BH-corrected)",
      align     = "llrrl")  %>%
  footnote(general = paste0(
    "Spearman correlations used due to non-normality of several variables. ",
    "p-values adjusted for multiple comparisons using Benjamini-Hochberg correction. ",
    "* = p < .05."
  ))
```


