---
title: "vo-dq"
author: "Iskra Todorova"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# REQUIRED PACKAGES
pkgs <- c("tidyverse", "ggplot2", "dplyr", "patchwork", "knitr", 
          "viridis", "DT", "kableExtra", "lme4", "emmeans", 
          "lmerTest", "ggdist", "tidyquant", "data.table")

# Install packages if not installed
installed_packages = pkgs %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(pkgs[!installed_packages])
}

# Load all packages
invisible(lapply(pkgs, library, character.only = TRUE))

# Display Session Info
sessionInfo()
```

## Introduction
This document performs the data quality pipeline for the Visual Oddball task.
Steps:
1.PD NA check & summary: Identifying missing pupil dilation data.
2.Trial Exclusion: Removing trials with > 50% missing data.
3.Gaze Offset Check: Excluding individual samples where gaze deviated more than 200px from the center.

### Data loading

```{r load-data}
# Paths – adjust to your project
home_path <- "S:/KJP_Studien"
data_path <- "/LOCUS_MENTAL/6_Versuchsdaten/visual_oddball/"

# Load processed files
df_et <- readRDS(paste0(home_path, data_path, "eyetracking_processed.rds"))
```

### PD NA Check & Summary
We analyze the missingness of the Pupil Dilation (pd) column at both the trial level and the participant level.

```{r pd-na-analysis}
# Per-trial NA summary
trial_na_summary <- df_et %>%
  group_by(id, trial_number) %>%
  summarise(
    total_samples = n(),
    na_count      = sum(is.na(pd)),
    na_prop       = mean(is.na(pd)),
    .groups = "drop"
  )

# Overall summary stats for NA proportions
cat("Summary of NA proportion per trial:")
summary(trial_na_summary$na_prop)

# Per-participant NA summary
participant_na_summary <- df_et %>%
  group_by(id) %>%
  summarise(
    total_samples = n(),
    na_count      = sum(is.na(pd)),
    na_prop       = mean(is.na(pd)),
    .groups = "drop"
  ) %>%
  arrange(desc(na_prop))

# Display table of participant missingness
DT::datatable(participant_na_summary, caption = "Missing PD Data per Participant")
```

### Trial Exclusion (> 50% Missing)
We remove trials that do not meet the data density threshold (50%).
```{r pd-exclusion}
na_threshold <- 0.50

high_na_trials <- trial_na_summary %>%
  filter(na_prop > na_threshold)

# Exclude flagged trials from the eye-tracking data
df <- df_et %>%
  anti_join(high_na_trials, by = c("id", "trial_number"))

# Exclusion summary
total_trials     <- nrow(trial_na_summary)
excluded_trials  <- nrow(high_na_trials)
remaining_trials <- total_trials - excluded_trials

# Create a summary data frame for reporting
excl_summary <- data.frame(
  Metric = c("Total Trials", "Excluded Trials", "Remaining Trials"),
  Count = c(total_trials, excluded_trials, remaining_trials),
  Percentage = c(100, (excluded_trials/total_trials)*100, (remaining_trials/total_trials)*100)
)

knitr::kable(excl_summary, digits = 2, caption = "NA Filtering Summary") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Gaze Offset Check (AOI = 200 px)
In this step, we average the gaze from both eyes and determine if the gaze remained within a ±200 px box from the center.

```{r gaze-preprocessing}
# Verify required gaze columns 
gaze_cols         <- c("left_gaze_x", "right_gaze_x", "left_gaze_y", "right_gaze_y")
missing_gaze_cols <- setdiff(gaze_cols, names(df))

if (length(missing_gaze_cols) > 0) {
  stop("Missing gaze columns: ", paste(missing_gaze_cols, collapse = ", "))
}

# Average gaze coordinates
df <- df %>%
  mutate(
    gaze_x_px = (left_gaze_x + right_gaze_x) / 2,
    gaze_y_px = (left_gaze_y + right_gaze_y) / 2
  )

# AOI classification (Screen center is 0,0)
aoi_threshold <- 200

df <- df %>%
  mutate(
    aoi_label      = ifelse(
      abs(gaze_x_px) <= aoi_threshold & abs(gaze_y_px) <= aoi_threshold,
      "AOI", "Offset"
    ),
    gaze_distance  = sqrt(gaze_x_px^2 + gaze_y_px^2), 
    gaze_deviated  = gaze_distance > aoi_threshold   
  )

# Report classification counts
aoi_counts <- table(df$aoi_label)
aoi_prop   <- round(prop.table(aoi_counts) * 100, 2)
print(aoi_counts)
print(aoi_prop)
```

#### Gaze Exclusion Summary

```{r gaze-exclusion -stats}
total_samples_before <- nrow(df)
df_aoi               <- df %>% filter(aoi_label == "AOI")
excluded_samples     <- total_samples_before - nrow(df_aoi)

gaze_summary <- data.frame(
  Metric = c("Total Samples", "Excluded (Offset)", "Remaining (AOI)"),
  Count = c(total_samples_before, excluded_samples, nrow(df_aoi)),
  Percentage = c(100, (excluded_samples/total_samples_before)*100, (nrow(df_aoi)/total_samples_before)*100)
)

knitr::kable(gaze_summary, digits = 2, caption = "Sample-level Gaze Exclusion") %>%
  kableExtra::kable_styling(full_width = F)
```

### Final Output Statistics

```{r final-report}
# Per-trial offset summary
trial_offset_summary <- df %>%
  group_by(id, trial_number) %>%
  summarise(
    total_samples   = n(),
    offset_samples  = sum(aoi_label == "Offset", na.rm = TRUE),
    offset_prop     = mean(aoi_label == "Offset", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(offset_prop))

cat("Final Clean Dataset Summary:\n")
cat("- Clean rows: ", nrow(df_aoi), "\n")
cat("- Participants: ", length(unique(df_aoi$id)), "\n")

# Show top 10 trials with highest offset
head(trial_offset_summary, 10)
```

