---
title: "Visual Search Task Data Quality"
author: "Iskra Todorova & Nico Bast"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# REQUIRED PACKAGES
pkgs <- c("tidyverse", "ggplot2", "dplyr", "patchwork", "knitr", "viridis", 
          "DT", "kableExtra", "data.table", "readxl", "psych")

# Install packages if not installed
installed_packages = pkgs %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(pkgs[!installed_packages])
}

# Load packages
invisible(lapply(pkgs, library, character.only = TRUE))
```

### Load Data


```{r load-data}
# PATHS (adjust to your project)
home_path <- "S:/KJP_Studien"
data_path <- "/LOCUS_MENTAL/6_Versuchsdaten/visual_search_task/"

# Load processed files
df_et <- readRDS(paste0(home_path, data_path, "eyetracking_processed.rds"))
df_trial <- readRDS(paste0(home_path, data_path, "trialdata_processed.rds"))
df_trial_hits_only <- readRDS(paste0(home_path, data_path, "eyetracking_trial_hits_only.rds"))
df_trial_hits <- readRDS(paste0(home_path, data_path, "eyetracking_trial_hits.rds"))

# Initial Cleaning
df_trial <- df_trial[is.na(df_trial$baseline_fixation_start_timestamp), ]
df_trial$trial_type <- ifelse(df_trial$auditory_cue == TRUE, "cued", "standard")
```

### Trial-Level Metrics
We examine the distributions of various trial phases (Trial Duration, Stimulus, ISI, and Beep phases) against their expected durations.

```{r trial-metrics}
# Constants
expected_trial_duration <- 3.5 
expected_stimulus_duration <- 1.5
expected_ISI_duration <- 1.5
expected_beep_phase_duration <- 0.4
expected_beep_duration <- 0.2

# Plot 1: Trial Duration
p1 <- ggplot(df_trial, aes(x = trial_duration)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~trial_type) +
  xlim(c(0,10)) +
  geom_vline(xintercept = expected_trial_duration, linetype = "dashed", color = "red", linewidth = 1) +
  labs(title = "Distribution of Trial Duration", x = "Duration (s)", y = "Count") +
  theme_minimal()

# Plot 2: Stimulus Duration
p2 <- ggplot(df_trial, aes(x = actual_visual_search_duration)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~trial_type) +
  geom_vline(xintercept = expected_stimulus_duration, linetype = "dashed", color = "red") +
  labs(title = "Distribution of Stimulus Duration", x = "Duration (s)", y = "Count") +
  theme_minimal()

# Plot 3: ISI Duration
p3 <- ggplot(df_trial, aes(x = ISI_actual_duration)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~trial_type) +
  xlim(c(0,10)) +
  geom_vline(xintercept = expected_ISI_duration, linetype = "dashed", color = "red") +
  labs(title = "Distribution of ISI Duration", x = "Duration (s)", y = "Count") +
  theme_minimal()

# Display plots using patchwork
(p1 / p2 / p3)
```

### Pupil Dilation (PD) Data Quality
We filter trials based on the proportion of missing Pupil Dilation data. Trials with more than 50% missing data are excluded.
```{r pd-quality}
na_threshold <- 0.5 

trial_na_summary <- df_et %>%
  group_by(id, trial_number) %>%
  summarise(
    total_samples = n(),
    na_count = sum(is.na(pd)),
    na_prop = mean(is.na(pd)),
    .groups = "drop"
  )

high_na_trials <- trial_na_summary %>% filter(na_prop > na_threshold)

# Exclude trials
df <- df_et %>% anti_join(high_na_trials, by = c("id", "trial_number"))

# Summary Table
total_trials <- nrow(trial_na_summary)
excluded_trials <- nrow(high_na_trials)
remaining_trials <- total_trials - excluded_trials

summary_df <- data.frame(
  Metric = c("Total Trials", "Excluded Trials", "Remaining Trials"),
  Count = c(total_trials, excluded_trials, remaining_trials),
  Percentage = c(100, (excluded_trials/total_trials)*100, (remaining_trials/total_trials)*100)
)

kable(summary_df, digits = 2, caption = "Exclusion Summary (Missing PD Data)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Gaze Position and AOI Analysis
Analysis of gaze coordinates to ensure participants remained within the monitor bounds and focused on the Area of Interest (AOI).
```{r gaze-aoi}
monitor_width <- 2560
monitor_height <- 1440
half_width <- monitor_width / 2
half_height <- monitor_height / 2

# Calculate average gaze
df$gaze_x_px <- (df$left_gaze_x + df$right_gaze_x) / 2
df$gaze_y_px <- (df$left_gaze_y + df$right_gaze_y) / 2

# Filter off-monitor
df_off_monitor <- df %>%
  filter(!is.na(gaze_x_px) & !is.na(gaze_y_px) & (abs(gaze_x_px) > half_width | abs(gaze_y_px) > half_height))
df <- anti_join(df, df_off_monitor)

# AOI logic (500px from center)
aoi_threshold <- 500
df_valid <- df %>%
  filter(!is.na(gaze_x_px) & !is.na(gaze_y_px)) %>%
  mutate(
    aoi_label = ifelse(abs(gaze_x_px) <= aoi_threshold & abs(gaze_y_px) <= aoi_threshold, "AOI", "Offset"),
    gaze_distance = sqrt(gaze_x_px^2 + gaze_y_px^2)
  )

df_aoi <- df_valid %>% filter(aoi_label == "AOI")

ggplot(df, aes(x = pd)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Pupil Dilation", x = "Pupil Dilation (mm)", y = "Frequency") +
  theme_minimal()
```

### Baseline Pupil Diameter
Statistical summary of baseline pupil diameter across participants and trial types.
```{r baseline-summary}
summary_stats <- df_aoi %>%
  group_by(id, trial_type) %>%
  summarise(
    mean = mean(mean_baseline_pd, na.rm = TRUE),
    sd = sd(mean_baseline_pd, na.rm = TRUE),
    n = sum(!is.na(mean_baseline_pd)),
    .groups = "drop"
  )

summary_stats %>%
  kable(caption = "Summary Statistics of Mean Baseline PD") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  scroll_box(height = "500px")

ggplot(df_aoi, aes(x = factor(id), y = mean_baseline_pd)) +
  geom_boxplot(outlier.color = "red", fill = "skyblue", alpha = 0.7) +
  facet_wrap(~trial_type, scales = "free_x") +
  labs(title = "Baseline Pupil Diameter by Participant", x = "Participant ID", y = "Mean Baseline PD") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Performance (Hits)
Evaluation of hit rates and response times (hit times) per condition.
```{r hits-analysis}
# Synchronize data frames
valid_pairs <- df_aoi %>% select(id, trial_number) %>% distinct()
df_trial_hits <- df_trial_hits %>% semi_join(valid_pairs, by = c("id", "trial_number"))
df_trial_hits_only <- df_trial_hits_only %>% semi_join(valid_pairs, by = c("id", "trial_number"))

# Hit Rate Summary
hit_summary <- df_trial_hits %>%
  group_by(id, trial_type) %>%
  summarise(hit_rate = mean(hit, na.rm = TRUE), .groups = "drop")

ggplot(hit_summary, aes(x = trial_type, y = hit_rate, fill = trial_type)) +
  geom_violin(alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "Hit Rate by Trial Type", x = "Trial Type", y = "Hit Rate") +
  theme_minimal()

# Hit Time Distribution
ggplot(df_trial_hits_only, aes(x = trial_type, y = hit_time, fill = trial_type)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Trial-Level Distribution of Hit Times", x = "Trial Type", y = "Hit Time (s)") +
  theme_minimal()
```

### Data Quality Checks & Export

```{r}
implausible_fast <- df_trial_hits_only[hit_time < 0.2]
implausible_slow <- df_trial_hits_only[hit_time > 1.5]

cat("Implausibly fast responses (<200ms):", nrow(implausible_fast), "\n")
cat("Implausibly slow responses (>1.5s):", nrow(implausible_slow), "\n")

# Save cleaned data
# saveRDS(df_aoi, file = paste0(home_path, data_path, "df_aoi_cleaned.rds"))
```

### Data 

```{r summary}
# 3. Calculate unique participant counts
count_aoi <- n_distinct(df_aoi$id)
count_hits <- n_distinct(df_trial_hits$id)
count_hits_only <- n_distinct(df_trial_hits_only$id)

# 4. Create a summary table to display results
filtering_summary <- data.frame(
  Dataset = c("df_aoi (Source)", "df_trial_hits (Filtered)", "df_trial_hits_only (Filtered)"),
  Unique_Participants = c(count_aoi, count_hits, count_hits_only)
)

knitr::kable(filtering_summary, caption = "Summary of Participant Counts after Filtering")

```
